### 2023-12-27 talk/aprm24.tex
draft of aprm24 talk

@@

\begin{center}
  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems with shared and unit-specific parameters}

%% ABSTRACT: We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes. Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population. Block particle filters are an effective tool for simulation-based likelihood evaluation for these systems, which are strongly dependent through time on a single unit and relatively weakly coupled between units. Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters. We introduce an iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units. We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal measles case report data in twenty towns. We discuss other applications, to cholera in Haiti and COVID-19 in China.

@@

Texas A\&M Statistics Colloquium\\
Apr 21, 2023

@@

Slides are at \url{https://ionides.github.io/talks/tamu23.pdf}

@@

Joint work with
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King

@@

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models? (Of particular interest are metapopulation models.)

@@

\item We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.

@@

\begin{frame}{Pre-vaccination weekly measles in England \& Wales}

  %% full weekly data now available:
  %% Korevaar, Hannah, C. Jessica Metcalf, and Bryan T. Grenfell. "Structure, space and size: competing drivers of variation in urban and rural measles transmission." Journal of the Royal Society Interface 17, no. 168 (2020): 20200010.
  %% https://royalsocietypublishing.org/doi/pdf/10.1098/rsif.2020.0010
  %% data at https://rs.figshare.com/collections/Supplementary_material_from_Structure_space_and_size_competing_drivers_of_variation_in_urban_and_rural_measles_transmission_/5036567/1

\vspace{-2mm}

\begin{center}
\includegraphics[width=8.4cm]{he10-data.pdf}

\vspace{-2mm}

\begin{frame}{Time series data, panel data \& spatiotemporal data}

  \bi
\item Looking at one unit (town) is \myemph{time series analysis}.

  \item Joint modeling of a few units (say, 2 or 3) is \myemph{multivariate time series analysis}.

\item Analysis of many time series, without consideration of dynamic interactions, is \myemph{panel data analysis}.

\item Allowing for coupling between units, we get \myemph{spatiotemporal analysis}, which in our context is \myemph{metapopulation analysis}.

%\vspace{3mm}

%Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?

@@

    \begin{itemize}
    \item  Iiterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:

@@

\begin{frame}{Fitting mechanistic models to panel data}

\bi
\item To investigate epidemiological dynamics in multiple cities, one can consider each city independently, perhaps modeling a background immigration rate of infections for each city.

\item \myemph{Decoupling} leads to panel data analysis, by assumption.

  \item Iterated filtering methods extend to panel data (\texttt{panelPomp}, Breto et al, {\it JASA}, 2019).

\item We must decide which parameters should be modeled as \myemph{shared} (i.e., the same for each unit) vs \myemph{unit-specific}.

%\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.

@@

\begin{frame}{Filtering $U$ units of a coupled measles SEIR model}

\begin{center}
\includegraphics[width=10cm]{mscale_loglik_plot-1.pdf}

\vspace{-4mm}

Simulated data using a gravity model with geography, demography and transmssion parameters corresponding to UK pre-vaccination measles (Ionides et al, {JASA}, 2021).

%\end{center}

@@

\begin{frame}
\frametitle{$U=40$ units for a coupled measles SEIR model}

\vspace{-2.7mm}

\begin{center}
\includegraphics[width=8cm]{slice_image_plot-1.pdf}
\end{center}

{\bf A}. Simulated Susceptible-Exposed-Infected-Recovered dynamics coupled with a gravity model (log of biweekly reported cases).

{\bf B}. Measles UK pre-vaccination case reports for the 40 largest cities.

%\end{center}

@@

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal transmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing exponent, $I^{\alpha}$.
\item Measurement fraction.
\item Population movement model: gravity constant.
\item Dynamic noise: process overdispersion.
\item Measurement overdispersion.

\ei
\vspace{3mm}
(Adding gravity coupling to the model of He et al, {\it JRSI}, 2010)
%\vspace{5mm}
%Plug-and-play methods let you add or subtract features

@@

\bi
\item BPF worked quickly, easily and reliably on our measles model filtering experiments.

@@

\frametitle{Measles likelihood slices for coupling parameter, $G$}

\vspace{-5mm}

\begin{columns}[T] % align columns
\begin{column}{.55\textwidth}
  \includegraphics[width=6cm]{slice_combined_plot-1.pdf}
\end{column}

\hspace{-5mm}
\begin{column}{.52\textwidth}

  \vspace{6mm}

  Simulating $15$ year of data from $U=40$ cities.
  Slice likelihood, varying $G$ with other parameters fixed.\\
   (Ionides et al, {\it JASA}, 2021).

  \vspace{3mm}

  {\bf A}. Evaluation using adapted bagged filter (ABF).

    \vspace{3mm}

  {\bf B}. Evaluation using block particle filter (BPF).

    \vspace{3mm}

  {\bf C}. Evaluation using EnKF.

  \vspace{3mm}
  A much easier task than computing a profile likelihood for real data.

\end{column}
\end{columns}

@@

\begin{frame}{Scalability needed for practical inference}

Large numbers of parameters
  \begin{itemize}
  \item  Initial conditions will typically have to be estimated for each unit.
  \item Various dynamic parameters and measurement parameters (e.g., reporting rate) may also need to be unit-specific to obtain a statistical fit to the data.
\item Working with hundreds of estimated parameters raises additional challenges on top of the high-dimensional coupled dynamics.
  \end{itemize}

A moderate numbers of spatial units is enough to open new possibilities.

  \begin{itemize}
  \item As soon as dimension exceeds capabilities of a particle filter (say, $U=5$) we are in situations where likelihood-based inference for general models has been inaccessible.

  \item 5-500 coupled units is our target problem size.

  \item Larger problems will need numerical approximations (e.g., EnKF).

@@

%%% A: 9 shared parameters; 4 unit-specific
%%% B: all unit-specific

\vspace{-6mm}

\begin{columns}[T] % align columns
\begin{column}{.75\textwidth}
  \includegraphics[width=6.8cm]{ibpf/search_diagnostics-1.pdf}
\end{column}

\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{10mm}

$\bullet$    IBPF applied to simulations and data.

$\bullet$    Multiple searches from random starting points.

$\bullet$    Top 25\% of searches are subsequently refined.

  \end{column}
\end{columns}

@@

\begin{columns}[T] % align columns
\hspace{-4mm}\begin{column}{.75\textwidth}
  \includegraphics[width=7.5cm]{ibpf/sim_search_boxplot-1.pdf}

    \includegraphics[width=7.5cm]{ibpf/data_search_boxplot-1.pdf}

  \end{column}

\hspace{-18mm}
\begin{column}{.4\textwidth}
  \vspace{5mm}

  {\bf IBPF for measles, 20 cities}\\

  $\bullet$ simulations (top panel)\\
  dash line: likelihood at truth

  $\bullet$ data (bottom panel)\\
  dash line: MLE for case C.

%$\bullet$    Multiple search refinements.

%    \vspace{5mm}

    (A) 9 shared parameters, $4\times 20$ unit-specific parameters

    (B) all parameters unit-specific

    (C) all unit-specific, with constant immigration instead of coupling

  \end{column}
\end{columns}

\begin{frame}{Inferences from the previous slide}

%%% Wilks' Theorem: improvement on truth is about (1/2) chisq(p)

\begin{itemize}
  \item
IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified

  \item
    When coupled by a gravity model, simulated data show evidence against a constant immigration model.

\item 
Data do not favor this gravity model over a constant immigration model.

%%% An underwhelming sounding result, but a necessary step to make progress,
%%% and previously we did not know how to do this.

\begin{frame}{He et al (2010) estimates. $r$ is rank correlation with $N_{1950}$}

\resizebox{12cm}{!}{
\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
\hline
  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
\hline
Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
\hline
Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
\hline
Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
\hline
Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
\hline
Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
\hline
Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
\hline
Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
\hline
Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
\hline
Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
\hline
Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
\hline
Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
\hline
Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
\hline
Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
\hline
Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
\hline
Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
\hline
Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
\hline
Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
\hline
Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
\hline
Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
\hline
London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
\hline
$\qquad r$ & 1 & 0.46 & \cellcolor{yellow!50} 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
\hline
\end{tabular}
}

\begin{frame}{Cholera in Haiti, 2010-2019}

  %%% Haiti and Dominican Republic share Hispaniola

% https://web.archive.org/web/20151106110552/http://www.ihsi.ht/pdf/projection/Estimat_PopTotal_18ans_Menag2015.pdf
%Map	Department	Capital	Area (km2)	Population	Density (Pop./km2)	Planning Region
%1	Artibonite	Gonaïves	4,987	1,727,524	350	Central
%2	Centre	        Hinche	        3,487	746,236	        210	Central
%3	Grand'Anse	Jérémie	        1,912	468,301	        240	South
%4	Nippes	        Miragoâne	1,268	342,525	        270	South
%5	Nord	        Cap-Haïtien	2,115	1,067,177	500	North
%6	Nord-Est	Fort-Liberté	1,623	393,967	        240	North
%7	Nord-Ouest	Port-de-Paix	2,103	728,807	        350	Central
%8	Ouest	        Port-au-Prince	4,983	4,029,705	810	West
%9	Sud-Est	        Jacmel	        2,034	632,601	        310	West
%10	Sud	        Les Cayes	2,654	774,976	        290	South

  % Hurricane Matthew: October 2016

\begin{columns}[T] % align columns

%\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{2mm}

  $\bullet$  820,000 reported cases

  $\bullet$ nearly 10,000 deaths

$\bullet$  Population of 10,000,000 in 10 d\'{e}partements

    $\bullet$  Lee et al (2020) developed 4 models to guide vaccination plans:

    %%% Elizabeth Lee, Johns Hopkins

    {\bf 1}. stochastic, national \\
    {\bf 2}. ODE, spatial \\
    {\bf 3}. stochastic, spatial\\
    {\bf 4}. agent based

\end{column}

\begin{column}{.62\textwidth}
  \includegraphics[width=7.5cm]{haiti/haiti.png}

  \begin{center}
    {  \tiny \url{https://doi.org/10.5194/nhess-20-471-2020}}
  \end{center}

\end{column}

\end{columns}

   $\bullet$ Wheeler et al (2023) continued the data analysis

\begin{frame}{Haiti cholera weekly reports 2010-2019 (log scale)}
\includegraphics[width=12cm]{haiti/Plot_Reported_Cases-1.pdf}
\end{frame}

\begin{frame}{Log-likelihoods of models and an ARMA benchmark}

  \begin{center}
\begin{tabular}{|l|rc|rc|rc|}
\hline
& Model~1& (p) & Model~2 & (p) & Model~3 & (p)
\\
\hline
Wheeler et al
& \cellcolor{yellow!50} -2731.3 & \cellcolor{yellow!50}(15)
& -21957.3 & (6)
& \cellcolor{yellow!50} -17850.4 & \cellcolor{yellow!50}(35)
\\
Lee et al
& -3030.9 & (20)
& -29367.4 & (6)
& -31840.8 & (29)
\\
Log-ARMA(2,1) \hspace{-2.5mm}
& -2803.7 & (4)  %% -0.5*5615.3+4
& \cellcolor{yellow!50} { -18027.0} & \cellcolor{yellow!50} (40)  %% -0.5*36133.9+40
& -18027.0 & (40)
\\
\hline
\end{tabular}
  \end{center}

  \vspace{3mm}

  Model 1 (stochastic, aggregated): \\
  We add process overdispersion, and beat the ARMA benchmark.

  Model 2 (deterministic, spatial): \\
  We use a log-normal measurement model. ODE models fall far short of the ARMA benchmark.

    Model 3 (stochastic, spatial): \\
    We fit the model using IBPF, and beat the ARMA benchmark.

\begin{frame}{Evidence for a decreasing transmission rate in Model 1}
\includegraphics[width=12cm]{haiti/Beta_trend_Figure-1.pdf}
\end{frame}

% From Li et al (2020) page S-14:
% Ruiyun Li, Imperial (Oslo as of 2021)
% According to official report, around 5 million people left Wuhan city before January 23. The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Inference using iterated EnKF.

    \vspace{5mm}
   \item
    The time interval predates Wuhan travel restrictions.
%% see README for R code
   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
   \item{} 0 cases reported for all cities during Jan 10--15.

\begin{frame}{Top: data. Middle: IBPF fit. Bottom: IEnKF fit} 
\includegraphics[width=12cm]{covid/3-panel.pdf}
\end{frame}

\begin{frame}{Technical issues with EnKF}

  \begin{itemize}
  \item EnKF is based on a continuous Gaussian approximation.
  \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
    \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
    \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
%    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF. The corresponding ``log-likelihood'' is not the log-likelihood of any model. In particular, it cannot validly be compared againt simple statistical benchmarks.
%  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
  \end{itemize}

\begin{frame}{Future work}

  \newcommand\futuresep{\vspace{5mm}}

  \begin{myitemize}

  \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.

    \futuresep

  \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.

%    \futuresep

 %   \item Questions and comments?

>>>>

%% for flow diagrams
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary {arrows.meta}
\usetikzlibrary{shapes.geometric}

@@

\begin{center}
  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems}

% ABSTRACT
% We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes. Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population. Block particle filters are an effective tool for simulation-based likelihood evaluation for these systems, which are strongly dependent through time on a single unit and relatively weakly coupled between units. Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters. We introduce a new iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units. We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal COVID-19 case report data in 373 cities.

@@

IMS Asia Pacific Rim Meeting\\
Jan 5, 2024

@@

Slides are at \url{https://ionides.github.io/talks/aprm24.pdf}

@@

Joint work with
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual

@@

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models? In particular, models for networks of interacting biological population dynamics.

@@

\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in \texttt{spatPomp}.

@@

@@

    \begin{itemize}
    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:

@@

@@

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Li et al (2020) used iterated EnKF for inference.
  \item The time interval covers the initial China lockdown.
  \item Here, we summarize our re-analysis of this model and data, recently posted on arXiv.
%% see README for R code
%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
%   \item{} 0 cases reported for all cities during Jan 10--15.

@@

\begin{frame}
  %%{An SEAIR metapopulation model for COVID-19}
\begin{center}
%%%%% SEAIR diagram
\resizebox{11cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
  bendy/.style={bend left=25},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (S) at (-0.5,0) [square] {S$_u$};
\node (E) at (2,0) [square] {E$_u$};
\node (A) at (4.5,0) [square] {A$_u$};
\node (I) at (4.5,-1.75) [square] {I$_u$};
\node (R) at (7,0) [square] {R$_u$};

\node (T1) at (-0.5,1.75) [travel] {T};
\node (T2) at (2,1.75) [travel] {T};
\node (T3) at (4.5,1.75) [travel] {T};

\node (Ca) at (4.5,-3.5) [report] {C$^a_u$};
\node (Cb) at (7,-3.5) [report] {C$^b_u$};
\node (C) at (9.5,-3.5) [report] {C$_u$};

\node (V1) at (2.5,-1.75) {};
\node (V2) at (3,-3.5) {};
\draw [->] (E) -- (Ca -| V2) -- (Ca);
\draw [->] (I -| V1) -- (I);

\draw [->] (S) -- (E);
\draw [->] (A) -- (R);

\draw [->] (Ca) -- (Cb);
\draw [->] (Cb) -- (C);

\draw [->] (I.east) -- (R);
\draw [->] (E) -- (A);

\draw [->, bendy] (S) to (T1);
\draw [->, bendy] (T1) to (S);

\draw [->, bendy] (E) to (T2);
\draw [->, bendy] (T2) to (E);

\draw [->, bendy] (A) to (T3);
\draw [->, bendy] (T3) to (A);

\end{tikzpicture}
}
\end{center}
\bi
\item 
  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
  \item 
An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
\item
  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
  \item 
Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
\ei

@@

\begin{frame}
  \begin{center}
    \includegraphics[height=9.5cm]{covid/panel_plot-1.png}
    \end{center}
\end{frame}

\begin{frame}
 \includegraphics[width=12cm]{covid/filter_tests10.png}

  \begin{itemize}
\item BPF is the best available filter for this COVID model.
\item Similar results hold for other models (Ionides et al, {\it JASA}, 2023).
  \end{itemize}

@@

\bi
\item BPF also worked quickly, easily and reliably on a measles metapopulation, as well as various toy benchmark problems (Ionides et al, {\it JASA}, 2023).

@@

@@

@@

\begin{frame}{Practical inference using IBPF}
  \bi
  \item to do

@@

%% \begin{frame}{Technical issues with EnKF}

%%   \begin{itemize}
%%   \item EnKF is based on a continuous Gaussian approximation.
%%   \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
%%     \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
%%     \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
%% %    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF. The corresponding ``log-likelihood'' is not the log-likelihood of any model. In particular, it cannot validly be compared againt simple statistical benchmarks.
%% %  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
%%   \end{itemize}

%%   \end{frame}

%% \begin{frame}{Future work}

%%   \newcommand\futuresep{\vspace{5mm}}

%%   \begin{myitemize}

%%   \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.

%%     \futuresep

%%   \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.

%% %    \futuresep

%%  %   \item Questions and comments?

%% \end{myitemize}

%% \end{frame}
